---
title: "Project - US Homicide Statistics Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

<h3>Load Libraries</h3>
```{r}
library(tidyverse)
library(ggplot2)
library(skimr)
library(scales)
library(wesanderson)
```


<h3>Load Dataset</h3>
```{r}
homicide_stats <- read_csv("HomicideStats.csv")
```


<h3>Summary of Data</h3>
```{r}
skim(homicide_stats)
```


```{r}
summary(homicide_stats)

```


<h3>Univariate Numerical - Crime Rate over the time period</h3>
```{r}
homicide_stats %>%
  ggplot(aes(x = Year))+
  geom_density(fill = "cyan", color = "black")+
  labs(title = "Count of Total criminal cases per year",
       x = "Year",
       y = "Count of criminal cases")

homicide_stats %>%
  group_by(Year) %>%
  mutate(Count = n()) %>%
  select(Year, Count) %>%
  distinct()

# homicide_stats %>%
#   ggplot(aes(x = Year, fill = State))+
#   geom_histogram()+
#   labs(title = "Count of Total criminal cases per year",
#        x = "Year",
#        y = "Count of criminal cases")

```


<h3>Univariate Categorical (three different years - min, median, max)</h3>
<h3>Count of criminal cases per State for different years</h3>
```{r}
homicide_stats %>%
  filter(Year == c(min(Year), median(Year), max(Year))) %>%
  ggplot(aes(x = State))+
  geom_bar(fill = "skyblue", color = "black")+
  facet_grid( ~ (Year))+
  coord_flip()+
  labs(title = "Count of criminal cases per State for different years",
       x = "State",
       y = "Count of criminal cases")

```


<h3>Univariate Categorical - Most Preferred Weapon from 1980-2014</h3>
```{r}
homicide_stats %>%
  ggplot(aes(x = Weapon, y = ..count..))+
  geom_bar(fill = "navyblue", color = "black")+
  coord_flip()+
  labs(title = "Most preferred Weapon",
       x = "Type of Weapon Used",
       y = "Count of weapon ")+
  scale_y_continuous(labels = comma)

```



<h3>Univariate Analysis Numerical - Which age group below 15 years is most 
victimised?</h3>
```{r}
homicide_stats_age15 <- homicide_stats %>%
  filter(`Victim Age` <= 15) %>%
  mutate(age_group = cut(`Victim Age`, breaks=c(0,5,10,15), include.lowest = T, labels = c("0-5 Years", "6-10 Years", "11-15 Years")))

homicide_stats_age15 <- homicide_stats_age15 %>%
  group_by(age_group, `Victim Sex`) %>%
  mutate(count = n(),
         `Concern Count` = cut(count, breaks = c(0,5000,10000,15000), include.lowest = T, labels = c("<5000","Between 5000 and 10000","More than 10000")))

#palette
pal <- c("<5000" = "yellow", 
         "Between 5000 and 10000" = "orange", 
         "More than 10000" = "red")

homicide_stats_age15 %>%
  filter(!is.na(age_group)) %>%
  ggplot(aes(x = age_group, fill = `Concern Count`))+
  geom_bar()+
  facet_grid( ~ `Victim Sex`)+
  labs(title = "Overall Analysis of Crime against Victims below 15 years",
       x = "Age Groups of Victims",
       y = "Total Count of Victims")+
  coord_flip()+
  scale_x_discrete(limits = rev(levels(homicide_stats_age15$age_group))) +
  scale_fill_manual(
    values = pal,
    limits = names(pal))
```


<h3>Univariate Categorical</h3>
<h3>Count of Criminal Cases per State for different decades</h3>
```{r}
#library(forcats)
homicide_stats_decade <- homicide_stats %>%
  mutate(decade = cut(Year, breaks = c(1980,1989,1999,2009,2014), include.lowest = T, labels = c("1980-1989","1990-1999","2000-2009","2010-2014")))

homicide_stats_decade_order <- homicide_stats_decade %>%
  select(State, decade) %>%
  group_by(State,decade) %>%
  mutate(Count = n()) %>%
  distinct() 

homicide_stats_decade_order %>%
  #mutate(Count = fct_reorder(State,desc(Count))) %>%
  ggplot(aes(x = reorder(State, Count), y = Count))+
  geom_bar(stat = "identity", fill = "cyan", color = "black")+
  coord_flip()+
  facet_grid( ~ decade)+
  labs(title = "Count of criminal cases per State for different decades",
       x = "State",
       y = "Count of criminal cases")


```


<h3>Victim Race Crime Analysis</h3>
```{r}
homicide_stats_victim <- homicide_stats %>%
  mutate(Total = n()) %>%
  group_by(`Victim Race`) %>%
  select(`Victim Race`, Total) %>%
  mutate(Count = n()) %>%
  mutate(Percentage = round(Count*100/Total,2)) %>%
  distinct()

# Gradient color
#grad <- wes_palette("Zissou1", 100, type = "continuous")

homicide_stats_victim %>%
  ggplot(aes(x = `Victim Race`, y = Percentage))+
  geom_bar(stat = "identity", aes(fill = Percentage, color = "red"))+
  geom_text(aes(label = Percentage))+
  labs(title = "Victim Race Crime Analysis",
       x = "Victim Race",
       y = "%age")+ 
   scale_x_discrete(expand = c(0, 0)) #+
   #scale_y_discrete(expand = c(0, 0)) + 
   #coord_equal()

```



<h3>Bivariate Analysis (Categorical+Categorical)</h3>
<h3>Victim Race vs Perpetrator Race Crime Analysis</h3>
```{r}
homicide_stats_race <- homicide_stats %>%
  mutate(Total = n()) %>%
  group_by(`Victim Race`,`Perpetrator Race`) %>%
  select(`Victim Race`,`Perpetrator Race`, Total) %>%
  mutate(Count = n()) %>%
  mutate(Percentage = round(Count*100/Total,2)) %>%
  distinct()

# Gradient color
grad <- wes_palette("Zissou1", 100, type = "continuous")

homicide_stats_race %>%
  ggplot(aes(x = `Victim Race`, y = `Perpetrator Race`))+
  geom_tile(aes(fill = Percentage, color = "red"))+
  geom_text(aes(label = Percentage))+
  labs(title = "Victim Race vs Perpetrator Race Crime Analysis",
       x = "Victim Race",
       y = "Perpetrator Race")+
  scale_fill_gradientn(colours = grad) + 
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) + 
  coord_equal()

```


<h3>Analysis of Crime Solving Rate of various Agencies</h3>
```{r}

homicide_stats_t <- homicide_stats %>%
  group_by(`Agency Type`,`Crime Solved`) %>%
  select(`Agency Type`,`Crime Solved`) %>%
  mutate(Count = n()) %>%
  distinct()


homicide_stats_t %>%
  ggplot(aes(x = `Agency Type`, y = Count, fill = `Crime Solved`))+
  geom_bar(stat = "identity")+
  coord_flip()+
  labs(title = "Analysis of Crime Solving Rate of various Agencies",
       x = "Agency Type",
       y = "Count of Criminal Cases")+
  scale_y_continuous(labels = comma)


```


<h3>Conclusion</h3>
1. On analysis, we have found out that the crime was prevalent in 1990s. Post that, we have decline in the crime rate.
2. From 1980 to 2014, California is the leading state when it comes to crime, which is followed by Texas.
3. Handgun is the most accessible weapon and thus, is used most in the criminal activities.
4. Below 15 years, most of the criminal cases are reported in case of children between 0-5 years of age for both males and females.
5. First two decades of the dataset, i.e., 1980-1989 and 1990-1999 have the highest crime rates as compared to rest of the years.
6. 49.72% of the criminal cases are reported where the victims belong to white race whereas 46.97% of the cases belong to the victims who are of black race.
7. 30.88% of the cases belong to the category where both the perpetrator and the victim belonged to the white race whereas 28.08% of the cases belong to the category where both are black.
8. Most of the cases are solved by Municipal Police. Most of the unsolved cases are also in the basket of Municipal Police.























